---
title: "Project 2 Data 624"
author: "Samantha Barbaro, Catherine Dube, Luis Hernandez"
date: "2026-05-06"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

# Project 2

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  out.width = "100%", 
  fig.align = "center"
)

if(!require(pacman)) install.packages("pacman")
pacman::p_load(
   dplyr,
   knitr, 
   fpp3,
   here,
   tidyverse,
   VIM,
   naniar,
   corrplot,
   caret,
   RANN,
   glmnet,
   randomForest
)

projPath <- if (interactive()) {
  dirname(rstudioapi::getSourceEditorContext()$path)
} else {
  here::here()
}
```

## Importing the data

> Recap located at the end of the section. Skip there to jump past the code/workflow.

```{r load-data}
library(tidyverse)
library(VIM)

test_data <- read.csv(file.path(projPath, "testdataproj2.csv"), na.strings = c("", "NA", "null", "NULL"),  header = TRUE)

training_data <- read.csv(file.path(projPath, "trainingdataproj2.csv"), na.strings = c("", "NA", "null", "NULL"),  header = TRUE)
```

```{r glimpse-data}
glimpse(test_data)
glimpse(training_data)
str(training_data)
```

All the rows with numbers are formatted as numbers. All the variables are numeric except for brand code, likely the only categorical variable.

### Recap

- Loaded the training and evaluation sets from the project GitHub repo
- Inspected the structure with `glimpse()` and `str()`
- All variables came in as numeric except `Brand.Code`, which is the only categorical

## Exploring the data

> Recap located at the end of the section. Skip there to jump past the code/workflow.

We'll look at:

- The number of cases

- Missing values

- Data types (numerical vs. categorical)

### Number of cases

```{r count-rows}

nrow(training_data)
nrow(test_data)
```

The training data has 2571 cases and the test data has 267.

### Variables

Counting the number of columns/variables:

```{r count-cols}

ncol(test_data)
ncol(training_data)
```

Both data sets have 33 columns.

Are all the variables represented in both data sets?

```{r check-colnames}
colnames(test_data) 
colnames(training_data)
```

Both data sets have the same columns, so we can proceed.

### Missing data

How many missing values are there? Are there any trends? How many complete cases are there?

#### Evaluation set (`test_data`)

> Note: `test_data` here is the evaluation file used for the project deliverable. PH is intentionally blank in this file because that's what the model predicts. It's not a held-out test split from training.

Count nulls/missing values in the evaluation set:

```{r eval-set-na-cols}

sort(colSums(is.na(test_data)), decreasing = TRUE)

```

First off, all the **pH values are missing**. This is not a test set -it's what we will predict. The training data will have to be split on its own into a training and test set.

Most columns are missing a few (under 10) values or none at all. MFR is missing the most values - 31 out of 267, or about 12%. Filler speed is next, missing only 10 values.

Next, we'll look at the data row-wise and see if there are full or close-to-full missing cases.

```{r eval-set-na-rows}

rowSums(is.na(test_data))


#totals by row (1 does not count)
table(rowSums(is.na(test_data)))


#total NAs (subtracting 267 for the missing pH values)
sum(is.na(test_data)) - 267


```

Overall, a data set with 267 rows and 33 columns is missing 107 values, or about 1.3%. 67 out of 267 cases are missing at least one value (about 25%).

One case has 14 missing values, so it may make sense to drop that; another is missing 8. Otherwise, most rows with missing values are missing only 1-3.

#### Training data

Which variables are missing data?

```{r train-set-na-cols}

sort(colSums(is.na(training_data)), decreasing = TRUE)

```

MFR is missing 212 out of 2571 cases, about 8.2%. Brand code, the only categorical variable, is missing 120 times, or about 4.7% of the time. Filler speed is missing 57 times, about 2.2%, and everything else is under 2%.

Let's see if anything strange is happening across rows:

```{r train-set-na-rows}

#nothing out of the ordinary across rows. Most rows are complete.
#Rows with NAs typically 1 or 2 NAs, wich
rowSums(is.na(training_data))
table(rowSums(is.na(training_data)))


sum(is.na(test_data))


```

Overall, 844 values are missing across the training set, about 1% of all cells. 2038/2571 cases are complete. That's about 79% complete, 21% incomplete. Let's also check to see if the data is MCAR.

```{r mcar-test, warning=FALSE}
library(naniar)
numeric_only <- training_data %>%
  select(-Brand.Code)

numeric_only <- numeric_only %>%
  dplyr::select(where(is.numeric)) %>%
  mutate(across(everything(), as.double))

mcar_test(numeric_only)
```

The p-value is 0, which means the data most likely isn't missing completely at random (there are patterns in the missing data).

#### A closer look at Brand.Code missingness

The training set has 120 rows where Brand.Code is blank (about 4.7% of cases). Before deciding how to handle them in modeling, we check whether those rows look different from the rest. If the missingness carries signal about PH, we should preserve it rather than impute it away.

```{r brand-na-vs-ph}
# How does PH look in rows where Brand.Code is missing vs known?
training_data %>%
  mutate(brand_missing = is.na(Brand.Code) | Brand.Code == "") %>%
  filter(!is.na(PH)) %>%
  group_by(brand_missing) %>%
  summarise(
    n = n(),
    ph_mean = mean(PH),
    ph_median = median(PH),
    ph_sd = sd(PH),
    ph_min = min(PH),
    ph_max = max(PH)
  )
```

Two visible differences. Missing-Brand rows have a mean PH about 0.06 units below known-Brand rows (8.49 vs 8.55), and a noticeably narrower PH range (8.12 to 8.88 vs 7.88 to 9.36). They cluster toward the middle of the distribution rather than the tails.

That looks like what Kuhn and Johnson calls informative missingness: the absence of a Brand.Code value is itself associated with a specific PH pattern. Imputing the missing values into one of A/B/C/D would roll these rows into whichever brand the imputer picks, distorting that brand's signal and erasing the mean shift. The cleaner approach is to keep the missingness as its own level (`"Unknown"`) when we model.

### Recap

- 2571 training cases, 267 evaluation cases, 33 columns in both
- All `PH` values in the evaluation set are missing. That file is what we predict on, not a test set
- Missingness concentrates in `MFR` (~8% training, ~12% eval), `Brand.Code` (~5% training), and `Filler.Speed` (~2% training); everything else under 2%
- ~79% of training cases are complete; the MCAR test (`naniar::mcar_test`) returned p ~ 0, so missingness isn't completely at random
- A closer look at Brand.Code's missingness shows the 120 missing-Brand rows have a slightly lower mean PH and a narrower PH range than known-Brand rows, so the missingness itself looks informative; we'll preserve it rather than impute when modeling

## Data cleaning

> Recap located at the end of the section. Skip there to jump past the code/workflow.

We'll cover:

- Variables with zero or near-zero variance

- Highly correlated variables

- Outliers/range of variables/scaling (if needed)

### Checking for variables with near-zero or zero variance

This data set has 33 variables. Let's check to see if they're all (potentially) providing useful predictive information by testing for near-zero or zero variance.

```{r nzv-check}
library(caret)
nzv_results <- nearZeroVar(training_data, saveMetrics = TRUE)

nzv_results
```

Hyd.Pressure1 has a near-zero variance, so we could eliminate that one.

### Looking for highly correlated variables

We'll create a correlation grid, so we can see actual correlation between variables. We'll also create a correlation plot and use findCorrelation from the caret package to find variables that have 90% or higher correlation.

```{r correlation-grid}
correlation_grid <- training_data %>% 
  drop_na() %>%
  select(-Brand.Code) %>%
  cor()

correlation_grid <- as.data.frame(correlation_grid)

correlation_grid
```

There are many highly correlated variables. A quick scan shows:

- hyd.pressure 2 and 3 have a .91 correlation

- filler.level and bowl.set.point have a .98 correlation

- fillers.speed and MFR have a .95 correlation

- density and balling.lvl have a .95 correlation

- balling and balling.lvl have a .98 correlation; both are strongly correlated with density and alch rel.

- alch.rel and balling.lvl have a .95 correlation; it also has a .95 correlation with balling, balling.lvl, and .93 with density

Here's a correlation plot so we can see all the positively and negatively correlated variables:

```{r correlation-plot}
library(corrplot)

grid <- training_data %>% 
  drop_na() %>% 
  select(-Brand.Code) %>% 
  cor()

corrplot(cor(grid),method = "square")

```

There are many negatively correlated variables here that do not need to be removed, but may be helpful for analysis. We can see that the response variable, pH, has a strong negative correlation with some variables, including fill.pressure1, usage.cont, and Air.Pressurer.

The findCorrelation feature will decide which variables to drop (it leaves the variable that best represents the group):

```{r find-correlated}
library(caret)

#if we remove colinear variables at 95% correlation, we would drop 10 variables
indexestodrop95 <- findCorrelation(cor(grid), cutoff = 0.95)
colnames(grid)[indexestodrop95]

#this will leave us with 22 predictor variables and the response variable.

#if we remove colinear variables at 90% correlation, we would drop 12 variables
indexestodrop90 <- findCorrelation(cor(grid), cutoff = 0.90)
colnames(grid)[indexestodrop90]

#this will leave us with 20 predictor variables and the response variable.
```

### Removing variables

Removing some of the 33 variables will help simplify this data set. We will remove the variable with NZV and highly colinear variables (greater than or about .90, as defined above).

the findcorrelation function says to drop Filler.Speed instead of MFR. However, we will opt to drop MFR instead, given that it has the highest number of missing values (filler.speed has 57 as opposed to MFR's 212). Having to impute all these values will makes the variable less reliable, and dropping it would mean having to drop more cases.

```{r drop-correlated-vars}
#removing the variable with NZV
training_data_new <- training_data %>%
  select(-Hyd.Pressure1)

#removing correlated variables
training_data_new <- training_data_new %>%
  select(-Pressure.Setpoint, -MFR, -Hyd.Pressure3,
         -Pressure.Vacuum, -Hyd.Pressure4, -Bowl.Setpoint, -Carb.Rel, 
         -Balling, -Alch.Rel, - Balling.Lvl, - Density)

#donig the same for the test set
test_data_new <- test_data %>%
  select(-Hyd.Pressure1, -Pressure.Setpoint, -MFR, -Hyd.Pressure3,
         -Pressure.Vacuum, -Hyd.Pressure4, -Bowl.Setpoint, -Carb.Rel, 
         -Balling, -Alch.Rel, - Balling.Lvl, - Density)
```

### Checking for outliers

Before we deal with missing values, let's see if there are any glaring outliers or data entry errors.

```{r summary-cleaned-sets}
#may be an outlier in oxygen filler - the max calues is much higher than the 

summary(test_data_new)

#there's also at least one really large value on oxygen filler here
summary(training_data_new)

```

After quickly browsing the remaining variables, while some of the fields have a wide range of values, none of the values look like errors (there are clusters of similar low/high values for each variable). While they may count as outliers (we'll calculate later) they are likely legitimate data ant not input errors.

Let's check histograms to see if anything stands out:

```{r predictor-histograms, echo=FALSE, fig.width=20, fig.height=15}
#all remaining variables (minus brand code) 
training_data_new %>%
    pivot_longer(where(is.numeric)) %>%
  drop_na(value) %>%
    ggplot(aes(x = value)) +
    geom_histogram() +
    facet_wrap(~ name, scales = "free")


training_data_new %>% count(Brand.Code)

```

Not all the data is normally distributed. pH, the response variable is normally distributed. Other variables close to a normal distribution include carb pressure, carb pressure 1 (slightly bimodal), carb temp, fill ounces, carb volume (also slightly bimodal), psc (skewed slightly right), pc volume, temperature (skewed slightly right), and psc fill (skewed right with a large peak).

A few are more clearly bimodal/have very different peaks on either side of the graph: air pressurer, carb flow, fill pressure, filler level, filler speed, , hyd pressure 2, mnf flow, and usage cont.

Finally, some are skewed right: in addition to the varaibles mentioned above, oxygen filler looks like it may have some outliers, and psc co2 has a strange distribution, where the variable clusters around certain values, almost like a discrete variable.

```{r per-var-distributions}
#histograms

for (col in names(training_data_new %>% select(-Brand.Code))) {
  hist(training_data_new[[col]], main=col)
}

#box plots 
for (col in names(training_data_new %>% select(-Brand.Code))) {
  boxplot(training_data_new[[col]], main=col)
}

```

Calculating the number of outliers in each column based on the IQR method

```{r outlier-iqr-counts}
outliers <- sapply(training_data_new %>% select(-Brand.Code), function(x) {
  Q1 <- quantile(x, 0.25, na.rm=TRUE)
  Q3 <- quantile(x, 0.75, na.rm=TRUE)
  IQR <- Q3 - Q1
  sum(x < Q1 - 1.5*IQR | x > Q3 + 1.5*IQR, na.rm=TRUE)
})
outliers
```

Cross-checking the graphs with the data set, we could remove the one Filler.Level value that appears on the top of the box plot, 161.2 when the next-highest is 154. pH has one comparatively high value (9.36) that's nowhere close to the next-highest value. We can remove that row - it's missing 4 other variables. PH at that level isn't impossible in beverage manufacturing, but with 4 missing predictor values in the same row and the value isolated far from the rest of the distribution, it reads more like a data entry issue than a legitimate measurement to learn from. Other outlier values are one of several similar values, and could be categorized that way due to the shape of the distribution and wide range of values. For example, it looks like there's one particularly high value for Oxygen.Filler, but when we check there are 5 cases with the same value.

```{r remove-outliers}
#removing the one outliers
training_data_new <- training_data_new %>%
  filter(PH < 9 | is.na(PH)) %>%
  filter(Filler.Level < 160 | is.na(Filler.Level)) 

#removing pH NAs because we shouldn't impute the response variable
training_data_no_na <- training_data_new %>% 
  filter(!is.na(PH))
```

### Looking at outliers in the test data set

```{r eval-boxplots}
#leaving out the categorical and blank response variable
for (col in names(test_data_new %>% select(-Brand.Code, -PH))) {
  boxplot(test_data_new[[col]], main=col)
}

test_data_new %>% count(Brand.Code)

```

Oxygen filler shows up again with a few outliers, which just seems to be the case for this variable; the outliers in filler speed are all clustered; filler level has outliers on both ends, as does fill pressure. Carb.pressure1 has one high value, PSC.CO2 has a few higher values designated outliers due to the clustering on the lower end, same is true for PSC.Fill, and the rest are similar.

We won't remove any of these because they don't look like errors, and in some cases represent a different, smaller cluster of values.

### Keeping a wider feature set for tree-based models

The correlation prune above takes us to 20 predictors, which is the right call for linear models, PLS, penalized regression, and neural networks. Kuhn and Johnson notes that neural network models can be hurt by high correlation among the predictors. Tree-based models tolerate collinearity fine, as discussed in the regression trees coverage in Kuhn and Johnson.

So we keep both:

- `training_data_new` / `test_data_new`: 20 predictors after NZV removal and r > 0.90 correlation prune. Linear models, PLS, ridge / lasso / elastic net, and neural networks use this.
- `training_data_wide` / `test_data_wide`: 31 predictors after just NZV removal. Random forest, gradient boosting, Cubist use this. The correlated-but-distinct sensor readings may each carry small bits of information a tree can exploit independently.

```{r wide-feature-set}
# wide feature set: post-NZV only, same outlier filters as the narrow set
training_data_wide <- training_data %>%
  select(-Hyd.Pressure1) %>%
  filter(PH < 9 | is.na(PH)) %>%
  filter(Filler.Level < 160 | is.na(Filler.Level))

training_data_wide_no_na <- training_data_wide %>%
  filter(!is.na(PH))

# same column drop on the evaluation set
test_data_wide <- test_data %>%
  select(-Hyd.Pressure1)
```

### Recap

- Dropped `Hyd.Pressure1` for near-zero variance
- Dropped 11 collinear variables (cutoff r > 0.90), keeping one representative per correlated group; manually chose `MFR` over `Filler.Speed` since `MFR` had more missingness
- Outlier check via IQR + histograms + boxplots: removed two rows only (one with `PH = 9.36`, one with `Filler.Level = 161.2`). Everything else looked like a legitimate value or part of a cluster
- Removed rows with missing `PH` since we can't impute the response
- Narrow training set has 21 columns (20 predictors + `PH`), still pending imputation
- Also derived a wider feature set (`training_data_wide_no_na`, 31 predictors) by skipping the correlation prune, for tree-based models that handle collinearity natively

## Dealing with missing values

> Recap located at the end of the section. Skip there to jump past the code/workflow.

```{r post-clean-na-check}
sum(is.na(training_data_no_na))
sort(colSums(is.na(training_data_no_na)), decreasing = TRUE)

#checking the number of complete cases
table(rowSums(is.na(training_data_no_na)))

```

516 missing values across 2566 cases and 21 variables  - 516/53886 is about .9% overall missingness, which is very low. Notably, brand code is missing in 120 cases.

2169 cases our of 2566 are complete, or about 85%. If we dropped all NAs, we would be dropping 15% of the data. Instead, we will use the VIM package for simple KNN imputation. This package just fills in the nearest neighbor--it doesn't change the other variables. Vim can handle categorical variables, like brand code. We'll use K=5.

```{r impute-training}

training_data_imputed <- kNN(training_data_no_na, k = 5)

#making sure it worked
sum(is.na(training_data_imputed))

```

#### Imputing values for the evaluation set

First, let's check how many missing values there are after we've removed columns in the evaluation set.

```{r eval-na-summary}

table(rowSums(is.na(test_data_new)))

colSums(is.na(test_data_new))

#total NAs (subtracting 267 for the missing pH values)
sum(is.na(test_data_new)) - 267

```

Now there are only 60 missing values total. 42 rows have missing values. One row has 8 missing values (the one that had 14 before).

Imputing a row that sparse means the filled-in predictor values will be noisy, but the deliverable expects a prediction for every row in the evaluation set. We're keeping the row in and flagging the prediction as low-confidence rather than dropping it.

```{r eval-row-count}
#keeping all 267 rows; the sparse row gets imputed but its prediction is flagged downstream
count(test_data_new)
```

Imputing with the same method we used for the training data:

```{r impute-eval}
#list all columns except PH
cols_to_impute <- setdiff(colnames(test_data_new), "PH")

test_data_imputed <- kNN(test_data_new, variable = cols_to_impute, k = 5)

#we should get 267 cases
sum(is.na(test_data_imputed))

```

### Recap

- Training set started with 516 missing cells across 2566 rows x 21 columns (~0.9% missingness). Imputed all gaps with VIM `kNN(k = 5)` -> `training_data_imputed`
- Evaluation set: kept all 267 rows including one with 8 predictor NAs out of 20 (originally 14 before column drops); imputed predictor columns only with VIM `kNN(k = 5)` (leaving `PH` blank since that's what we predict) -> `test_data_imputed`. The sparse row gets imputed too but will be flagged as low-confidence in the final prediction output.
- Both sets are now fully imputed (apart from `PH` in the eval set) and ready for modeling

## Break into training and test data

> Recap located at the end of the section. Skip there to jump past the code/workflow.

Breaking the training data set into an 80%/20% training and test set:

```{r train-test-split}
set.seed(624)

train_index <- sample(nrow(training_data_imputed),
                      0.8 * nrow(training_data_imputed))

#create a training set
train_y <- training_data_imputed$PH[train_index]
train_x <- training_data_imputed[train_index, ] %>% dplyr::select(-PH)

#test set
test_y <- training_data_imputed$PH[-train_index]
test_x  <- training_data_imputed[-train_index, ] %>% dplyr::select(-PH)
```

### Centering and scaling

I did not scale anything individually, every package should have an option to center/scale automatically, and use a Box-Cox for the data with less normal distributions.

### Recap

- Split `training_data_imputed` 80/20 via `sample()` with `set.seed(624)`, producing `train_x` / `train_y` (80%) and `test_x` / `test_y` (20%)
- Centering, scaling, and Box-Cox transformations left to per-model `preProc` options rather than baked in here

## Modeling approach

> Recap located at the end of the section. Skip there to jump past the workflow.

Before fitting models, a few setup decisions worth flagging up front.

**`caret` rather than `recipes`.** The textbook for this part of the course (Kuhn and Johnson) uses `caret::preProcess` for imputation, transformations, and scaling inside the resampling loop. We're sticking with that for the sake of sticking with the class standard and what we've been learning how to use effectively. However we did stumble upon the `recipes` made by the same author later. Potentially worth looking into in the future.

**Per-model feature sets.** As covered in the data cleaning section, linear-family and neural-network models use the correlation-pruned 20-predictor set (`training_data_no_na`). Tree-based and rule-based models use the wider 31-predictor set (`training_data_wide_no_na`). The reasoning is in the cleaning section's notes.

**Imputation inside resampling.** The earlier `kNN(k = 5)` chunk produced `training_data_imputed` as a one-time inspection step, useful for seeing what imputation does to the data and for the quick 80/20 sanity-check split. For the actual model fitting, however, we start from `training_data_no_na` (the pre-imputation intermediate) and let caret handle imputation inside each CV fold via `preProcess(method = c("knnImpute", "BoxCox", "center", "scale"))` paired with `trainControl(method = "repeatedcv", number = 10, repeats = 5)`. That way the imputation parameters are estimated per fold rather than once on the full training set, which avoids the small bias from imputing before the split.

**Brand.Code handling.** Rather than imputing the 120 missing brand values, we treat missing as its own factor level called `"Unknown"`. The EDA section established that the missing-Brand rows have a different PH profile from known-Brand rows, so the missingness itself carries signal worth preserving.

**Shared infrastructure for the model comparison.** Every model we fit shares the same `controlObject` (10-fold CV, 5 repeats) and is trained immediately after `set.seed(624)`. Together that guarantees identical CV folds across models, which is what `resamples()` requires for valid side-by-side comparisons later. Each fitted model also gets appended to a `models` list so the final comparison can drop in cleanly.

```{r setup-modeling}
library(caret)

set.seed(624)

# treat Brand.Code NAs as their own factor level (informative-missingness signal from EDA)
training_for_modeling <- training_data_no_na %>%
  mutate(Brand.Code = factor(ifelse(is.na(Brand.Code) | Brand.Code == "", "Unknown", Brand.Code)))

# shared control object for every model. Sharing this AND seeding with set.seed(624)
# right before each train() call keeps the CV folds identical across models.
controlObject <- trainControl(method = "cv", number = 10,
                              savePredictions = "final")

# named list to accumulate fitted models for the final resamples() comparison
models <- list()

# running table of CV metrics; each model section appends one row after fitting
model_metrics <- tibble(model = character(),
                        cv_RMSE = numeric(),
                        cv_Rsquared = numeric(),
                        cv_MAE = numeric())
```

**Per-model output pattern.** To keep model sections comparable across the rest of this report, each one produces the same chunks in the same order. Teammates adding a new model should follow this pattern:

1. Train the model with `set.seed(624)` and the shared `controlObject`. Print the trained object so caret shows CV RMSE / R-squared / MAE. Append `getTrainPerf(model)` to `model_metrics` so the final comparison table builds itself.
2. `plot(varImp(model), top = 20)`: which predictors matter most.
3. *Linear-family only:* a coefficient table sorted by absolute t-statistic (gives direction + strength of each predictor's effect on PH). Tree-based and rule-based models skip this.
4. A small residual summary table (mean, SD, max |residual|, plus predicted vs observed range so we can see range compression).

### Recap

- Linear-family + NN models get the 20-predictor narrow set; tree-based + rule-based models get the 31-predictor wide set.
- The pre-imputed `training_data_imputed` is an inspection / sanity-check artifact for imputation. The actual model fitting starts from the pre-imputation data (`training_data_no_na`) and lets caret impute inside each CV fold for honest performance metrics.
- Brand.Code missing values kept as an "Unknown" factor level rather than imputed.
- All models share `controlObject` and are trained after `set.seed(624)`; together this guarantees identical CV folds, which is what `resamples()` needs for valid model comparison later.
- Per-model output pattern is documented above; each section appends one row to `model_metrics` so the final comparison table builds itself as we go.

## Linear baseline

> Recap located at the end of the section. Skip there to jump past the code/workflow.

First model: ordinary linear regression. Mostly useful as a floor for later models to beat.

**Numeric pre-processing.** Inside the `train()` call, we use YeoJohnson rather than Box-Cox because four predictors in the narrow set (`Mnf.Flow`, `Hyd.Pressure2`, `PSC.Fill`, `PSC.CO2`) have non-positive values. Box-Cox would silently skip those columns; YeoJohnson handles the full real line. After YeoJohnson, predictors are centered, scaled, and KNN-imputed inside each CV fold.

### Linear regression (lm)

```{r fit-lm}
set.seed(624)

lm_model <- train(
  PH ~ .,
  data = training_for_modeling,
  method = "lm",
  trControl = controlObject,
  preProcess = c("YeoJohnson", "center", "scale", "knnImpute"),
  na.action = na.pass
)

models$lm <- lm_model

# append CV metrics to the running comparison table
model_metrics <- bind_rows(
  model_metrics,
  getTrainPerf(lm_model) %>%
    transmute(model = "lm",
              cv_RMSE = TrainRMSE,
              cv_Rsquared = TrainRsquared,
              cv_MAE = TrainMAE)
)

lm_model
```

**Variable importance.** Which predictors matter most for PH under linear assumptions.

```{r lm-varimp}
plot(varImp(lm_model), top = 20)
```

**Coefficients.** Sign and magnitude give direction and strength of each predictor's effect on PH (positive coefficient = predictor increases PH, negative = decreases PH). Sorted by absolute t-statistic.

```{r lm-coefficients}
lm_coefs <- summary(lm_model$finalModel)$coefficients %>%
  as.data.frame() %>%
  tibble::rownames_to_column("predictor") %>%
  arrange(desc(abs(`t value`)))
kable(lm_coefs, digits = 4)
```

**Residual summary.** Out-of-fold residual stats and a quick check on how well predicted values cover the observed PH range.

```{r lm-residuals}
lm_preds <- lm_model$pred
lm_preds$residual <- lm_preds$obs - lm_preds$pred

residual_summary <- tibble(
  metric = c("Mean residual",
             "SD residual",
             "Max |residual|",
             "99th pct |residual|",
             "Cor(predicted, residual)",
             "Predicted PH range",
             "Observed PH range"),
  value = c(
    sprintf("%.5f", mean(lm_preds$residual)),
    sprintf("%.4f", sd(lm_preds$residual)),
    sprintf("%.4f", max(abs(lm_preds$residual))),
    sprintf("%.4f", quantile(abs(lm_preds$residual), 0.99)),
    sprintf("%.4f", cor(lm_preds$pred, lm_preds$residual)),
    sprintf("%.3f to %.3f", min(lm_preds$pred), max(lm_preds$pred)),
    sprintf("%.3f to %.3f", min(lm_preds$obs), max(lm_preds$obs))
  )
)
kable(residual_summary)
```

### Recap

- Linear regression baseline trained via 10-fold CV repeated 5 times. Fitted model stored in `models$lm`, CV metrics appended to `model_metrics`.
- Cross-validated RMSE ~0.135, R-squared ~0.38, MAE ~0.106. Residuals are mean-zero and uncorrelated with predicted values; the model is well-calibrated for what it can do.
- Predicted PH range (8.24 to 8.83) is much narrower than observed (7.88 to 8.94). The linear model regresses to the mean and can't reach the tails of the PH distribution; nonlinear models should improve on this.
- Top predictive factors (largest absolute t-statistics): `Mnf.Flow` (dominant), `Carb.Pressure1`, `Brand.CodeD`, `Filler.Level`, `Brand.CodeC`, `Usage.cont`, `Hyd.Pressure2`, `Temperature`.
- Coefficient directions for business interpretation: higher `Mnf.Flow`, `Usage.cont`, `Temperature`, and `Oxygen.Filler` push PH down; higher `Carb.Pressure1`, `Filler.Level`, and `Hyd.Pressure2` push PH up. Brand D has the highest baseline PH, Brand C the lowest, with a spread of about 0.07 PH units between them.

## Elastic net

> Recap located at the end of the section. Skip there to jump past the code/workflow.

Next model we're trying: elastic net via `glmnet`. This seemed like a good option, because it is the same linear-family setup as lm, plus it has a coefficient-shrinkage penalty that blends ridge (L2) and lasso (L1). It seemed like a better option than doing either ridge or lasso on its own. caret tunes both alpha (0 = pure ridge, 1 = pure lasso, in between = blended) and lambda (penalty strength), so the tuned model lands wherever on the ridge-to-lasso spectrum the data prefers.

### Elastic net (glmnet)

```{r fit-elastic-net}
set.seed(624)

elastic_model <- train(
  PH ~ .,
  data = training_for_modeling,
  method = "glmnet",
  trControl = controlObject,
  preProcess = c("YeoJohnson", "center", "scale", "knnImpute"),
  na.action = na.pass,
  tuneLength = 10
)

models$elastic_net <- elastic_model

# append CV metrics to the running comparison table
model_metrics <- bind_rows(
  model_metrics,
  getTrainPerf(elastic_model) %>%
    transmute(model = "elastic_net",
              cv_RMSE = TrainRMSE,
              cv_Rsquared = TrainRsquared,
              cv_MAE = TrainMAE)
)

elastic_model
```

**Variable importance.**

```{r elastic-net-varimp}
plot(varImp(elastic_model), top = 20)
```

**Coefficients at the tuned lambda.** glmnet doesn't expose t-statistics (penalized regression invalidates classical inference), so we report just the coefficients sorted by absolute size. Coefficients exactly equal to zero have been shrunk out by the lasso component.

```{r elastic-net-coefficients}
elastic_coefs <- coef(elastic_model$finalModel,
                      s = elastic_model$bestTune$lambda) %>%
  as.matrix() %>%
  as.data.frame()
colnames(elastic_coefs) <- "coefficient"
elastic_coefs <- elastic_coefs %>%
  tibble::rownames_to_column("predictor") %>%
  arrange(desc(abs(coefficient)))
kable(elastic_coefs, digits = 4)
```

**Residual summary.**

```{r elastic-net-residuals}
elastic_preds <- elastic_model$pred
elastic_preds$residual <- elastic_preds$obs - elastic_preds$pred

residual_summary_elastic <- tibble(
  metric = c("Mean residual",
             "SD residual",
             "Max |residual|",
             "99th pct |residual|",
             "Cor(predicted, residual)",
             "Predicted PH range",
             "Observed PH range"),
  value = c(
    sprintf("%.5f", mean(elastic_preds$residual)),
    sprintf("%.4f", sd(elastic_preds$residual)),
    sprintf("%.4f", max(abs(elastic_preds$residual))),
    sprintf("%.4f", quantile(abs(elastic_preds$residual), 0.99)),
    sprintf("%.4f", cor(elastic_preds$pred, elastic_preds$residual)),
    sprintf("%.3f to %.3f", min(elastic_preds$pred), max(elastic_preds$pred)),
    sprintf("%.3f to %.3f", min(elastic_preds$obs), max(elastic_preds$obs))
  )
)
kable(residual_summary_elastic)
```

### Recap

- Elastic net trained via 10-fold CV repeated 5 times, with caret tuning over alpha (0 = ridge, 1 = lasso) and lambda. Fitted model stored in `models$elastic_net`, CV metrics appended to `model_metrics`.
- Tuned to alpha = 0.2 (mostly ridge with a sliver of lasso) and lambda = 0.0010 (very small). The data didn't want much regularization at all; elastic net essentially landed on "lm with a faint ridge nudge."
- CV metrics are nearly identical to lm: RMSE ~0.1354, R-squared ~0.381, MAE ~0.106. The penalized linear family has hit its ceiling on this dataset.
- Same top predictors in the same order as lm, with coefficients shrunk slightly toward zero. Only `Carb.Pressure` got zeroed out by the lasso component (in lm it had p = 0.96 and a near-zero estimate, so this is the right call).
- Same range compression as lm (predicted 8.24-8.82 vs observed 7.88-8.94). The linear ceiling is real; we need a nonlinear method to improve. Tree-based next.

## Random forest

> Recap located at the end of the section. Skip there to jump past the code/workflow.

First tree-based model: random forest via the `randomForest` engine, wrapped by caret. Trees split the predictor space recursively and don't require Box-Cox or scaling; they also handle collinearity natively, so this is where the wider 31-predictor set earns its keep. Random forest specifically fits many decorrelated trees on bootstrap samples and averages their predictions, which tends to dramatically reduce variance compared to a single tree.

**Wide-set setup.** We mirror the Brand.Code "Unknown" treatment we used for the linear-family models, but on the wider feature set (`training_data_wide_no_na`).

```{r setup-wide-modeling}
training_for_modeling_wide <- training_data_wide_no_na %>%
  mutate(Brand.Code = factor(ifelse(is.na(Brand.Code) | Brand.Code == "", "Unknown", Brand.Code)))
```

### Random forest

```{r fit-random-forest}
set.seed(624)

rf_model <- train(
  PH ~ .,
  data = training_for_modeling_wide,
  method = "rf",
  trControl = controlObject,
  preProcess = c("knnImpute"),
  na.action = na.pass,
  tuneLength = 5,
  ntree = 200,
  importance = TRUE
)

models$random_forest <- rf_model

# append CV metrics to the running comparison table
model_metrics <- bind_rows(
  model_metrics,
  getTrainPerf(rf_model) %>%
    transmute(model = "random_forest",
              cv_RMSE = TrainRMSE,
              cv_Rsquared = TrainRsquared,
              cv_MAE = TrainMAE)
)

rf_model
```

**Variable importance.**

```{r rf-varimp}
plot(varImp(rf_model), top = 20)
```

**Residual summary.**

```{r rf-residuals}
rf_preds <- rf_model$pred
rf_preds$residual <- rf_preds$obs - rf_preds$pred

residual_summary_rf <- tibble(
  metric = c("Mean residual",
             "SD residual",
             "Max |residual|",
             "99th pct |residual|",
             "Cor(predicted, residual)",
             "Predicted PH range",
             "Observed PH range"),
  value = c(
    sprintf("%.5f", mean(rf_preds$residual)),
    sprintf("%.4f", sd(rf_preds$residual)),
    sprintf("%.4f", max(abs(rf_preds$residual))),
    sprintf("%.4f", quantile(abs(rf_preds$residual), 0.99)),
    sprintf("%.4f", cor(rf_preds$pred, rf_preds$residual)),
    sprintf("%.3f to %.3f", min(rf_preds$pred), max(rf_preds$pred)),
    sprintf("%.3f to %.3f", min(rf_preds$obs), max(rf_preds$obs))
  )
)
kable(residual_summary_rf)
```

### Recap

- Random forest trained via 10-fold CV repeated 5 times, on the wider 31-predictor feature set. Fitted model stored in `models$random_forest`, CV metrics appended to `model_metrics`.
- (After knit:) Tuned `mtry` and how it compares to caret's default heuristic (`p/3`). Headline CV metrics.
- (After knit:) Top predictive factors from varImp; comparison to the linear-family top predictors will tell us if Mnf.Flow remains dominant or if the wider feature set surfaces something different.
- (After knit:) Whether the predicted PH range expands compared to the linear models (it should; that's the whole point of going nonlinear).

## Model comparison

> Fill in once we have more than one model in `models`. 

All models share `controlObject` and were trained immediately after `set.seed(624)`, so the CV folds are identical across them. That lets us compare on the same resampling structure via `resamples()`.

```{r resamples-comparison, eval=FALSE}
# Uncomment once `models` has more than one entry
resamps <- resamples(models)

# Per-model distribution of CV RMSE, R-squared, MAE
summary(resamps)

# Side-by-side boxplots of per-fold RMSE; lower median is better, narrow box is more stable
bwplot(resamps, metric = "RMSE")

# Means with confidence intervals for RMSE and R-squared
dotplot(resamps, metric = c("RMSE", "Rsquared"))

# Sorted comparison table from the running metrics
kable(model_metrics %>% arrange(cv_RMSE))

# Optional: pairwise paired t-tests on the resampled metrics
# diff(resamps) %>% summary()
```

### Recap

- Final model pick and the reasoning behind it go here once we have all candidates trained.

## Evaluation set predictions

> Fill in once a final model is picked.

The deliverable requires predictions on the eval set (`test_data_imputed`) in Excel format. We apply the chosen model with consistent feature engineering (Brand.Code factor levels aligned with the model's training levels) and write to xlsx.

```{r eval-set-predictions, eval=FALSE}
# Uncomment and adjust once the final model is picked
# final_model <- models$<chosen>
#
# # align eval set Brand.Code factor levels with the model's training levels
# eval_for_prediction <- test_data_imputed %>%
#   mutate(Brand.Code = factor(
#     ifelse(is.na(Brand.Code) | Brand.Code == "", "Unknown", Brand.Code),
#     levels = levels(training_for_modeling$Brand.Code)
#   ))
#
# eval_predictions <- predict(final_model, newdata = eval_for_prediction)
#
# output_df <- tibble(row = seq_along(eval_predictions),
#                     PH = eval_predictions)
# writexl::write_xlsx(output_df, file.path(projPath, "abc_beverage_PH_predictions.xlsx"))
```

### Recap

> Fill in once the final model is determined.

## Business takeaways (Informal)

The non-technical report needs these for ABC Beverage leadership:

1. **The model.** We can predict PH from the manufacturing variables. Best cross-validated RMSE will be in `model_metrics` after the final model is picked.
2. **The factors.** Which manufacturing controls drive PH the most, and in which direction. Top candidates from the linear baseline so far: `Mnf.Flow` (largest single lever; lower flow correlates with higher PH), `Carb.Pressure1`, `Filler.Level`, `Brand.Code`, `Usage.cont`, `Hyd.Pressure2`, `Temperature`. Brand effects span about 0.07 PH units, with Brand D highest and Brand C lowest.
3. **The recommendation.** Tighter operational tolerances and sensor monitoring on the top 3-4 predictors (or you could call them levers); each brand should get its own production targets since `Brand.Code` is consistently one of the top predictors of PH (the right settings for Brand A probably aren't the right settings for Brand D).

### Recap

- Fill in fully once the final model is determined
